Active learning using region-based sampling

### 2024-07-31 COLT/response.tex

\documentclass[12pt]{colt2023} %% Anonymized submission
\usepackage{times}
\begin{document}
\title{}
This is a response to reviewers 2, 3, 4 (we haven't received Review 1).

\vspace{.1in}

\noindent
{\bf Reviewer 2:}

\noindent
\emph{Finiteness of the metric space?} We are given a finite set of points $X$ to begin with; the set of balls on these points, $\mathcal{B}$, is effectively finite.

\noindent
\emph{Connection to paper of Balsubramani et al?} We were indeed inspired by this earlier work; but that paper was not about active learning and doesn't directly yield a good active learning method. For instance, that paper assesses the status of a point $x$ using balls \emph{centered} at $x$, whereas for us it is crucial to use balls \emph{containing} $x$ but possibly centered elsewhere. Our three key challenges (section 1.2) are all specific to active learning and demand new approaches.

\vspace{.1in}
\noindent
{\bf Reviewer 3:}

\noindent
\emph{Questions about the algorithm:} A \emph{background query} is drawn at random from the entire data space while a \emph{focused query} is drawn from the current region of uncertainty (top of section 2.3). In Figure 3, each data point $x$ has an associated random variable $T_x$ drawn uniformly on $[0,1]$. The data points are ordered according to these values. The goal of this technical trick is to ensure that queries for any ball are drawn independently at random.

\noindent
\emph{Instantiations of Theorem 4:} Indeed, the theorem is pretty abstract. We give three concrete instantiations: two one-dimensional scenarios and a higher dimensional scenario with Massart noise and a curvature condition on the boundary (sec 4.3). It's also easy to give an instantiation under smoothness and Tsybakov noise (see below for Reviewer 4), which we plan to add in.

\noindent
\emph{Constants in Theorem 5:} Say the fraction of the data with $\eta$ value in $[0,0.4)$, $[0.4, 0.5)$, $(0.5, 0.6]$, and $(0.6,1]$ is $40\%, 10\%, 10\%, 40\%$ respectively. Then $n^- = n^+ = 0.4n$ and $r^- = r^+ = 0.1n$.

\noindent
\emph{Underlying assumptions?} Our \emph{algorithm} is pretty general, unlike many active learning procedures that make sense only under smoothness conditions, for instance. But our \emph{bounds} do need (smoothness or other) assumptions, as is inevitable in nonparametric estimation.

\noindent
\emph{More discussion:} We agree that the theorems and results need more discussion. We will figure out how to squeeze this in!

\vspace{.1in}
\noindent
{\bf Reviewer 4:}

\noindent
\emph{More discussion of results:} Some intuition behind the algorithm is in the introduction. The theorems are another matter, and certainly warrant more discussion. We will add this somehow!

\noindent
\emph{Theorem 9:} In Theorem 9, the rate of convergence is controlled by two parameters: $\sigma, \xi$ (the other constants are either inherent in the problem, like the number of points $n$ or data dimension $d$, or are absolute constants, like $C$). Both $\sigma$ and $\xi$ are exponents in analogs of the Tsybakov margin condition; the difference is that $\xi$ ignores all points whose $\eta$ value lies in $[1/2-\gamma, 1/2+\gamma]$, the range in which any prediction is acceptable.

We will also add another instantiation of Theorem 9, using a strong density assumption (as before), curvature condition on boundary (as before), Holder-smoothness (with exponent $\alpha$) and Tsybakov margin (with exponent $\beta$). In this case, it is easy to show that $\sigma = \xi = \alpha\beta$. We will add this to the main paper or the appendix, depending on space constraints.

\noindent
\emph{Related work:} A discussion of related work appears in Section 1.4, but this focuses on \emph{algorithmic strategies} rather than \emph{bounds}. One that is directly comparable, but is a special case, is Castro-Nowak (as mentioned at the bottom of page 12, we get similar bounds). But other work is hard to compare since the assumptions are different and specialized. For instance, Minsker has a special smoothness condition, Kpotufe et al have a probabilistic Lipschitz condition, etc. It is hard to bridge these gaps.

\end{document}

### 2024-07-31 COLT/colt.tex
Initial Overleaf Import

\documentclass[anon,12pt]{colt2022} % Anonymized submission
%\documentclass[final,12pt]{colt2022} % Include author names

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

\usepackage{amsfonts}

\def\R{{\mathbb{R}}}
\def\pr{{\textrm Pr}}
\def\E{{\mathbb E}}
\def\X{{\mathcal X}}
\def\Y{{\mathcal Y}}
\def\H{{\mathcal H}}
\def\G{{\mathcal G}}
\def\B{{\mathcal B}}
\def\yh{{\widehat{y}}}
\def\bias{{\rm bias}}
\def\supp{{\rm supp}}
\def\sign{{\rm sign}}
\def\PL{{\mbox{\rm PL}}}
\newcommand{\predict}{\mathbf{Predict}}
\newcommand{\learn}{\mathbf{Learn}}
\newcommand{\D}{{\cal D}}

\newcommand\bigsubset[1][1.19]{%
   \mathrel{\vcenter{\hbox{\scalebox{#1}{$\subset$}}}}}

\newcommand{\comment}[3]{{\color{#1} {\bf #2 :} #3}}
%\newcommand{\comment}[3]{}  % suppress comments
\newcommand{\sanjoy}[1]{\comment{red}{Sanjoy}{#1}}
\newcommand{\yoav}[1]{\comment{blue}{Yoav}{#1}}

@@

\title[Active learning using region-based sampling]{Active learning using region-based sampling}
\usepackage{times}
% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
\coltauthor{\Name{Sanjoy Dasgupta} \Email{dasgupta@eng.ucsd.edu}\and
\Name{Yoav Freund} \Email{yfreund@eng.ucsd.edu}\\
\addr UC San Diego}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

% Authors with different addresses:
%\coltauthor{%
% \Name{Author Name1} \Email{xyz@sample.com}\\
% \addr Address 1
% \AND
% \Name{Author Name2} \Email{xyz@sample.com}\\
% \addr Address 2%
%}

@@

\begin{abstract}%
Abstract goes here.
\end{abstract}

\begin{keywords}%
List of keywords
\end{keywords}

@@

\section{Setup}

We assume that examples are drawn from a fixed but unknown
distribution $\D$ over $\X \times \Y$, where $\X$ is a metric space and
$\Y = \{-1,+1\}$~\footnote{In a later section we define the notion of
  neighborhoods more generally to allow, for example, diadic
  partitions}
Given $x \in \X$, the label $y \in \Y$ is drawn
according to the conditional probability function
$$ \eta(x) = \E[Y | X=x] = \pr(Y=1|X=x) - \pr(Y=-1|X=x).$$
The \emph{Bayes-optimal classifier} on $\X$ is $g^*(x) = \sign(\eta(x))$.

The task we wish to solve is \emph{finite population annotation}. We
are given a finite dataset $X \subset \X$ of $n$ points.
We are given the ability to query the label of any point $x \in X$. The
answer to the query: $a(x) \in \{-1,+1\}$ is a binary random variable
drawn according to $\eta(x)$. In particular, multiple queries on the
same point can recieve different answers.

Our goal is to design an active learning algorithm that converges to the Bayes
optimal $g^*(x)$ and does so with a much smaller number of queries
than a passive algorithm.

We want a procedure that chooses the next point to query. It will be applied repeatedly and then stopped at an arbitrary time, whereupon labels $y(x)$ must be provided for all $x \in X$, including those that were queried. We would like the inferred labels $y(x)$ to approach the Bayes-optimal labels $g^*(x)$ as quickly as possible, ideally much faster than would occur if the query points were chosen uniformly at random.

Specifically, we will be judged on the number of errors on points $x \in X$ with $|\eta(x)| > \gamma$:
$$ \sum_{x \in X} {\bf 1}(y(x) \neq g^*(x),\, |\eta(x)| > \gamma) .$$
The procedure is allowed to fail with probability $\delta$, to account for sampling error.

\section{Active learning algorithm}

\subsection{Sampling regions}

The sampling is organized around a collection $\B$ of measurable subsets of $\X$. These are the atomic sets on which we assess label bias and we call them ``balls''.

\begin{enumerate}

\item[(a)] \emph{Points within a ball.} For any $B \in \B$, let $X_B$ denote the points in $X$ that can be used to assess the bias of $B$. This is simply $X_B = X \cap B$ if $\B$ is a predefined set of balls.

\item[(b)] \emph{Levels of sampling.} We group balls by the number of points that they contain: we put $B$ at {\it level} $\ell$ if
$$ \frac{n}{2^{\ell + 1}} \leq |X_B| < \frac{n}{2^\ell} .$$
Here $\ell$ ranges from $0$ (consisting of balls that contain at least half the data points) to roughly $\lg n$ (containing a single point).

Let $\B_\ell$ consist of all balls in $\B$ that are at level $\ell$. Thus $\B_0, \B_1, \ldots$ is a partition of $\B$, with $\B_0$ consisting of very large balls and $\B_1, \B_2, \ldots$ consisting of successively smaller balls.

\item[(c)] \emph{The neighborhood of a point.} For any
  $x \in \X$, and level $\ell$.
  $N(\ell,x)$ is the collection of balls from $\B_\ell$ that
  contain $x$.
\item[(d)] We say that the neighborhood $N(x,\ell)$ is {\em covered}
  if there are $k$ query points in each $B \in N(x,\ell)$.

\end{enumerate}

\subsection{Bias estimates and provisional labels}

The active learning algorithm uses label-queries to assess the bias of balls $B \in \B$. In turn, these are used to estimate the labels of individual points. All these estimates are organized by sampling level.

\begin{enumerate}

\item[(a)] We take the \emph{bias of a ball} $B \in \B$ to be
$\eta_X(B) = \mbox{average}\{\eta(x): x \in X_B \}$.
At any given time, the algorithm assigns $B$ a qualitative bias estimate,
$$ \yh(B) = 
\left\{
\begin{array}{cl}
+1 & \mbox{significant $+$ bias} \\
-1 & \mbox{significant $-$ bias} \\
0 & \mbox{no significant bias} \\
\bot & \mbox{not yet available}
\end{array}
\right.
$$
The option $\bot$ is used until sufficiently many points in $X_B$ have been labeled: the required number, $k$, is roughly $1/\gamma^2$ (recall that $\gamma$ is the smallest bias that needs to be detected). Once $k$ labels are available, $\yh(B)$ is set to a value in $\{-1,0,+1\}$ and remains fixed thereafter. It will satisfy a basic correctness guarantee (Definition~\ref{def:accurate-bias-estimate}) based on large deviation theory.

\item[(b)] \emph{Possible labels for a point.} Pick any $x$ and any level $\ell$. Once bias-estimates are available for all balls $B \in \B_\ell(x)$, the possible labels for $x$ at level $\ell$, denoted $\PL_\ell(x) \subset \{-1,0,+1\}$, are defined as follows:
\begin{gather}
+1 \in \PL_\ell(x) \Longleftrightarrow \mbox{there exists $B \in
  \B_\ell(x)$ with $\yh(B) = +1$} \notag \\
0 \in \PL_\ell(x) \Longleftrightarrow \mbox{there exists $B \in \B_\ell(x)$ with $\yh(B) = 0$} \notag \\
-1 \in \PL_\ell(x) \Longleftrightarrow \mbox{there exists $B \in \B_\ell(x)$ with $\yh(B) = -1$}
\label{eq:PL}
\end{gather}
The label of point $x$ at level $\ell$ is a value $\yh_\ell(x) \in \{-1,+1,0,!,\bot\}$. It is initially $\bot$, but once bias-estimates are available for all $B \in \B_\ell(x)$, it is set to a value in $\{-1,+1,0,!\}$ and remains fixed thereafter:
\begin{equation}
\yh_\ell(x) = 
\left\{
  \begin{array}{cl}
+1   & \mbox{if $+1 \in \PL_\ell(x)$ and $-1 \notin PL_\ell(x)$} \\
-1   & \mbox{if $+1 \notin \PL_\ell(x)$ and $-1 \in PL_\ell(x)$} \\
!    & \mbox{if $+1 \in \PL_\ell(x)$ and $-1 \in PL_\ell(x)$} \\
0    & \mbox{if $\PL_\ell(x)=\{0\}$} \\
\bot & \mbox{if $\PL_\ell(x)=\emptyset$}
\end{array}
\right.
\label{eq:provisional-label}
\end{equation}
A point $x$ for which $\yh_\ell(x)=!$ is said to have a ``conflict at
level $\ell$''.

\end{enumerate}

@@

\begin{figure}[h!]
\framebox{
\begin{minipage}[t]{5.9in}
\begin{itemize}
\item{\bf Initialize:} set $\ell=0$, cover all of the balls in $\B_0$.
\item   $\predict(x)$
  \begin{itemize}
  \item Predict with $\yh_\ell(x)$
  \end{itemize}
  \item
  $\learn$  Increment sampling level\\
  Repeat while there is an $x$ that has a conflict at level $0\leq i
  \leq \ell$ and $N(x,i+1)$ is not fully covered. 
  \begin{itemize}
  \item Cover $N(x,i+1)$ using $m$ queries
  \item Query $m$ points uniformly from the domain (background
    sampling)
  \item Update $\yh_{i+1}(y)$ For all $y$.
  \end{itemize}
  Let $\ell+1 \to \ell$
\end{itemize}
\end{minipage}}
\caption{The active learning algorithm. In each iteration, (at most) one focused query and one background query are made.}
\label{alg:main}
\end{figure}

>>>>

\documentclass[anon,12pt]{colt2024}

\usepackage{enumitem}
\usepackage{multicol}

\def\R{{\mathbb{R}}}
\def\pr{{\rm Pr}}
\def\E{{\mathbb E}}
\def\X{{\mathcal X}}
\def\Y{{\mathcal Y}}
\def\H{{\mathcal H}}
\def\G{{\mathcal G}}
\def\B{{\mathcal B}}
\def\yh{{\widehat{y}}}
\def\bias{{\rm bias}}
\def\supp{{\rm supp}}
\def\dist{{\rm dist}}
\def\sign{{\rm sign}}
\def\vol{{\rm vol}}
\def\PL{{\mbox{\rm PL}}}

\newtheorem{thm}{Theorem}
%\newtheorem{lemma}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
%\newtheorem{claim}[thm]{Claim}
\newtheorem{defn}[thm]{Definition}
\newtheorem{assump}{Assumption}
%\newenvironment{proof}{\noindent {\sc Proof:}}{$\Box$ \medskip}

\newcommand{\comment}[2]{ {\bf #1 :} #2}
\newcommand{\yoav}[1]{\comment{Yoav}{\em #1}}
\newcommand{\sanjoy}[1]{\comment{blue}{Sanjoy}{#1}}

@@

%\setlength{\leftmargini}{0.5cm}
%\setlength{\leftmarginii}{0.25cm}
%\setlength{\leftmarginiii}{0.25cm}

\title[Active learning using region-based sampling]{Active learning using region-based sampling}

\usepackage{times}

\coltauthor{
\Name{Sanjoy Dasgupta} \Email{dasgupta@eng.ucsd.edu}\and
\Name{Yoav Freund} \Email{yfreund@eng.ucsd.edu}\\
\addr University of California, San Diego}

@@

\begin{abstract}%
We present a general-purpose active learning scheme that maintains a collection of neighborhoods of different sizes and uses label queries to identify those with a strong bias towards one particular label. When two such neighborhoods intersect and have different labels, the region of overlap is treated as a ``known unknown'' and is targeted in future active queries. We give label complexity bounds for this method that do not rely on assumptions about the data, and we instantiate them in several cases of interest.
\end{abstract}

\begin{keywords}
Active learning, nonparametric estimation.
\end{keywords}

@@

In \emph{active learning}, the starting point is a data set whose labels are hidden but can be obtained individually for a price. The goal is to label the data set, or to find a good classifier, at low cost.

We consider a formulation in which we have an arbitrary collection of points $X = \{x_1, \ldots, x_n\}$ from some space $\X$. We can request the label of any of these points $x$, in which case we get a value $y \in \{-1, +1\}$ with conditional distribution
$ \eta(x) = \E[y | x] = \pr(y=1|x) - \pr(y=-1|x) $
for some unknown function $\eta: \X \to [-1,1]$. The \emph{Bayes-optimal} label for $x$, which we shall denote $g^*(x)$, is $-1$ if $\eta(x) < 0$ and $+1$ if $\eta(x) > 0$; either label can be used if $\eta(x) = 0$.
%$$ g^*(x) = \left\{
%\begin{array}{cl}
%-1 & \mbox{if $\eta(x) < 0$} \\
%+1 & \mbox{if $\eta(x) > 0$} \\
%\mbox{either} & \mbox{if $\eta(x)= 0$}
%\end{array} \right.
%$$
We wish to find the Bayes-optimal labels for the given data points $X$.

More precisely, at the outset we have the data $X$ and parameters $0 < \gamma, \delta < 1$. We want a procedure that chooses the next point, or batch of points, to query. This procedure will be applied repeatedly and can be stopped at any time, whereupon labels $\yh(x)$ must be provided for all $x \in X$, including those that were queried. Ideally, these will be the Bayes-optimal labels $g^*(x)$; however, we will only be judged on points with non-negligible bias, $|\eta(x)| \geq \gamma$. The number of mistakes is:
$$ \sum_{x \in X} {\bf 1} \big(\yh(x) \neq g^*(x),\ |\eta(x)| \geq \gamma \big) .$$
The overall procedure is allowed to fail with probability $\delta$, to account for sampling error.

\subsection{Approach and challenges}

In our setting there is no underlying assumption about $g^*$, for instance that it follows a linear model. Thus we adopt a nonparametric approach.

We begin with a large collection of \emph{neighborhoods}, $\B$, of varying sizes (Fig.~\ref{fig:challenges}(a)); we will refer to these as \emph{balls}. We use label queries to assign \emph{biases} $\yh(B) \in \{-1,0,+1\}$ to balls $B \in \B$, beginning with the largest balls. A bias of $+1$ means that the average $\eta(x)$ in $B$ is significantly positive; bias $-1$ means it is significantly negative; and bias $0$ means the average $\eta$ is close to 0. As time goes on, we get biases for progressively smaller balls, as needed. The final label assigned to any point $x$ is based on the biases of balls containing $x$.

\begin{figure}
\begin{center}
\includegraphics[width=5in]{ideas.pdf}
\end{center}
\caption{(a) A collection of balls $\B$. Ball $B$ has bias $\yh(B) \in \{-1,0,+1\}$. (b) If one ball has bias $+1$ and the other $-1$, points in their intersection are targeted for active querying. (c) Biases of progressively smaller balls containing $x$ might flip sign, leading to \emph{mind changes} about the label of $x$.}
\label{fig:challenges}
\end{figure}

%, me and to use label queries to estimate the signs associated with these regions. Ultimately we hope to cover $\X$ by a patchwork of neighborhoods with $\eta(x)$ values of uniform sign. Without strong smoothness conditions, these neighborhoods could be of very different sizes. The most beneficial for active learning are neighborhoods that contain a lot of $X$, but to get good coverage we might have to include smaller neighborhoods. To address this heterogeneity, we group the regions $\B$ by size. We begin by estimating the signs of the largest of them, and then move on to progressively smaller neighborhoods as the need arises. When the label budget runs out, the process is stopped, and provisional labels are assigned to the individual points in $X$.

There are three key challenges to be addressed. The first is deciding \emph{where to query}. Initial random querying tells us the biases of the largest balls, which gives some information about the decision boundary. In what parts of $\X$ might it be helpful to know biases of smaller balls, warranting further label queries? Our approach is simple: if a ball $B$ of bias $+1$ intersects a ball $B'$ of bias $-1$ then the intersection $B \cap B'$ is a ``known unknown'' and is targeted for active querying  (Fig.~\ref{fig:challenges}(b)).

This differs from the more conventional strategy of actively querying neighborhoods $B$ whose average $\eta$-value is close to zero. Such $B$ fall into two categories: (1) the $\eta$ values are close to zero throughout $B$, and (2) $B$ consists of two sub-regions, one strongly positive and the other strongly negative. In the first case, there is little merit in querying further. But in the second case, there is a lot to be gained. Our approach allows these two cases to be distinguished.

%To distinguish these two cases, we use a collection of neighborhoods that are \emph{overlapping}. For instance, we might take $\B$ to be \emph{all} balls in $\X$, which is effectively a finite collection once the given data points $X$ are taken into account. Case (2) can then be detected: we think of a point $x$ as being in the uncertainty region if it is contained in a neighborhood $B$ that is strongly positive as well as being in a neighborhood $B'$ that is strongly negative. Such points are ``known unknowns'', and these are the targets of our active querying.

A second challenge is that the sign of $\eta(\cdot)$ can change in an arbitrarily small neighborhood of any given point: the biases of balls containing $x$, say $B_1 \supset B_2 \supset \cdots$, might change sign a few times. At first, given only the bias of $B_1$, we might think $x$ has label $+1$. When we later estimate the bias of $B_2$, we could change our mind to $-1$. And then $+1$, and so on (Fig.~\ref{fig:challenges}(c)).

What this means is that we can \emph{never} be sure of having correctly determined the label of $x$. Thus in addition to focused (active) querying, we also do background sampling of the entire space to pick up on possible mind changes. For simplicity, we make one background query per focused query.

The third challenge is \emph{managing sampling} of overlapping balls. We are interested in detecting points, and thus neighborhoods, whose $\eta$-values are either $> \gamma$ or $<-\gamma$. This suggests querying $k \approx 1/\gamma^2$ points at random from each ball. Now, suppose we have queried this many points from ball $B$ and later want to query points from a different ball $B'$ that overlaps $B$. How can we reuse queries already made in $B \cap B'$?

\emph{Poisson sampling} provides a clean solution. Rather than choosing $k$ points at random from $B$, we pick each point in $B$ with probability (roughly) $k/|B \cap X|$, independently. We implement this by assigning each $x \in X$ a uniform-random value $T_x \in [0,1]$, at the outset. When sampling $B$, we choose to query $x$ if $T_x \leq k/|B \cap X|$. And when it comes time to sample a different $B'$ that also contains $x$, we choose $x$ if $T_x \leq k/|B' \cap X|$; if its label has already been obtained, we are able to reuse it. In this way, the random querying of overlapping balls is seamlessly managed.

These three challenges go beyond earlier work in active learning, which was able to avoid problems like mind-changes by making smoothness assumptions on $\eta$. By tackling all three of them, we are able to give a general-purpose active learning scheme.

\subsection{Results}

%Our algorithm does not require any assumption on the input distribution. In the worst case the number of queries is twice the number of queries of a passive algorithm. We show that under several natural conditions the query complexity of our algorithm is logarithmic in the error.

Section~\ref{sec:algorithm} presents the active learning algorithm and Section~\ref{sec:discrete} analyzes it in generality, taking $X$ to be an arbitrary point-set and allowing any $\eta$ function. We identify two \emph{critical levels} for any $x \in X$: two scales (of neighborhood sizes) that control how many queries are sufficient for $x$ to be correctly labeled (Theorem~\ref{thm:label-complexity}). We instantiate these bounds in canonical one-dimensional settings (Theorems~\ref{thm:oned-massart} and \ref{thm:oned-monotonic}) to get label complexities logarithmic in $|X|$.

Section~\ref{sec:continuous} considers the statistical setting where $X$ is drawn from an underlying distribution $\mu$ on $\X$. We give rates of convergence under distributional conditions (Theorem~\ref{thm:label-complexity-specific}), and instantiate them under common assumptions from learning theory and computational geometry.

\subsection{Related work}

Existing works on nonparametric active learning can mostly be grouped by overall principle: either (1) they seek to obtain the Bayes-optimal labels of a few well-positioned points, and then propagate these to the rest of the space \citep{DNZ15,H17,ASU18} or (2) they estimate the biases (positive or negative) of entire regions at a time \citep{DH08,M12}. In this paper, we follow the second strategy because the only reliable and general-purpose way to assess the bias of an individual point, $\mbox{sign}(\eta(x))$, is to query that point repeatedly; absent smoothness assumptions, the sign can change abruptly in an arbitrarily small neighborhood around $x$. On the other hand, the bias of a region $B$---the sign of the average $\eta$ value in $B$---is easy to determine given a reasonable number of points from that region.

Early results of \cite{CN08} established upper and lower bounds on label complexity in situations where the Bayes-optimal boundary is of a simple form: a single threshold for one-dimensional data or a ``smooth boundary fragment'' in higher dimension.

An algorithm for active learning based on hierarchical sampling was given by \cite{DH08} and was analyzed under smoothness conditions by \cite{KUB15}. The idea is to begin with a hierarchical clustering of $X$, and to then use queries to discover a pruning of this tree whose leaf-clusters are almost-pure in their labels. The method is not well-suited to situations with significant noise levels. Another approach using dyadic partitions was given by \cite{M12} and analyzed under commonly-used smoothness, margin, and density assumptions---namely, that $\eta$ is Holder-smooth, the fraction of points with $|\eta(x)| \leq t$ is some polynomial in $t$, and the marginal density is close to uniform---along with an additional smoothness requirement on $\eta$.

A different strategy using nearest neighbors was explored by \cite{ASU18}. Their idea was to choose an appropriate scale $s$, find an $s$-cover of $X$, estimate the Bayes-optimal label for each point in this cover by querying its neighbors, and then use these cleanly-labeled points for 1-nearest neighbor classification. A somewhat more general approach was given by \cite{H17} and studied under the usual smoothness, margin, and density conditions, with resulting rates of convergence comparable to those found by \cite{M12}.

Finally, \cite{DNZ15} suggested a graph-based method for active learning based on adaptively looking for the cut in the graph corresponding to the correct decision boundary. Their assumptions are based on properties of this cut and are not easily comparable with earlier work.

The idea of predicting with balls of multiple sizes is a central theme of \cite{BDFM19}, which predicts at a point $x$ by choosing a ball centered at $x$ that is as small as possible while having a clear bias towards one label. That work does not consider active learning, which demands considerable further innovation. For instance, to assess the uncertainty at a point $x$, we need to look at balls containing $x$, not just those centered at $x$ (Fig.~\ref{fig:challenges}(b)).

\section{The active learning algorithm}
\label{sec:algorithm}

Recall that we have a collection of points $X$ from a space $\X$. The label of any $x \in X$ can be requested and the value returned will be either $-1$ or $+1$, according to the conditional probability function $\eta(x) = \E[y|x]$. We wish to assign Bayes-optimal labels to all points in $X$ with $|\eta(x)| \geq \gamma$.

\subsection{A collection of sampling regions}

In our active learning algorithm, sampling is organized around a collection $\B$ of subsets of $\X$. These are the atomic sets on which we assess label bias and we call them \emph{balls} or \emph{neighborhoods}. There are no formal requirements on $\B$---for instance, the so-called balls can be of arbitrary shape---but the intention is that each $x \in X$ is contained in balls of multiple sizes, including one that is so small as to exclude the rest of $X$.

For any ball $B \in \B$, let $X_B = X \cap B$ be a shorthand for the data points that lie in it.
%\footnote{Since $X_B$ is used to assess the bias of $B$, we need to exclude points that cannot be considered random samples from $B$. For instance, consider these two alternatives: (1) $\B$ consists of a pre-defined set of balls. (2) $\B$ consists of balls defined by pairs of points in $X$, with each $x,x' \in X$ yielding the ball $B(x,\|x-x'\|)$. In the first case, all points $X \cap B$ are random draws from $B$. In the second case, we need to exclude the two points $x,x'$ that define the ball.} 
We group balls into levels by the number of points they contain. We put $B$ at {\it level} $\ell \geq 0$ if
\begin{equation}
\frac{n}{2^{\ell + 1}} \leq |X_B| < \frac{n}{2^\ell} .
\label{eq:sampling-level}
\end{equation}
Let $\B_\ell$ consist of all balls in $\B$ that are at level $\ell$. Thus $\B_0, \B_1, \ldots$ is a partition of $\B$, with $\B_0$ consisting of highly-populated balls and subsequent $\B_1, \B_2, \ldots$ consisting of successively smaller balls. We will use $\B_{\geq \ell}$ to denote all balls at levels $\ell$ or greater, and likewise $\B_{> \ell}$, $\B_{\leq \ell}$, and so on.

Balls in lower levels contain more points, and thus their biases (average $\eta$ values) are easier to estimate. Our sampling algorithm goes from easier to harder levels.

For any $x \in \X$, let $\B(x) \subset \B$ denote the collection of balls that contain $x$ and can thus be used in determining $x$'s label. We again partition these balls by sampling-level, so that $\B_\ell(x) = \B(x) \cap \B_\ell$. 
%We will also refer to $\B_\ell(x)$ as the {\em level $\ell$ neighborhood of $x$}.

\subsection{Estimating the biases of balls}

The bias of a ball $B \in \B$ is defined as the average $\eta$ value in it,
$$ \eta_X(B) = \mbox{average}\{\eta(x): x \in X_B \} .$$
We estimate these biases using label queries and assign each ball a \emph{qualitative} bias estimate,
$$ \yh(B) = 
\left\{
\begin{array}{cl}
+1 & \mbox{significant $+$ bias} \\
-1 & \mbox{significant $-$ bias} \\
0 & \mbox{no significant bias} \\
\bot & \mbox{not yet available}
\end{array}
\right.
$$

The option $\bot$ is used until enough points in $X_B$ have been queried: the required number is $k = O(1/\gamma^2)$ since $\gamma$ is the smallest bias that needs to be detected. Once this many labels are available, $\yh(B)$ is set to a value in $\{-1,0,+1\}$ and remains fixed thereafter. These bias estimates will with high probability be seen to satisfy the following guarantee.
\begin{defn}
For any $B \in \B$, bias estimate $\yh(B) \in \{+1,-1,0\}$ is \emph{$\gamma$-accurate} if:
\begin{itemize}
\item $\yh(B) = +1 \implies \eta_X(B) > 0$
\item $\yh(B) = -1 \implies \eta_X(B) < 0$
\item $\yh(B) = 0 \implies |\eta_X(B)| < \gamma$
\end{itemize}
\label{def:accurate-bias-estimate}
\end{defn}

\subsection{Assigning labels to points}

The label of any point is inferred from the biases of balls containing it.

Pick any $x \in X$ and any level $\ell$. Once qualitative bias estimates $\yh(B)$ are available for all balls at levels $\leq \ell$ that contain $x$ (that is, $\B_{\leq \ell}(x)$), we use these to infer a level-$\ell$ label for $x$. To begin with, the set of \emph{possible} labels for $x$ at level $\ell$, denoted $\PL_\ell(x) \subset \{-1,+1\}$, is defined thus:
\begin{itemize}[leftmargin=0.5cm]
\item $\PL_\ell(x)$ includes $+1$ if there exists a \emph{minimal} ball $B \in \B_{\leq \ell}(x)$ (with no other $B' \in \B_{\leq \ell}(x)$ strictly contained in it) for which $\yh(B) = +1$.
\item $\PL_\ell(x)$ includes $-1$ under a symmetrical condition.
\end{itemize}
This is made precise in Eqn~(\ref{eq:PL}) in Fig.~\ref{alg:main}. The label-estimate for $x$ at level $\ell$, denoted $\yh_\ell(x)$, is $+1$ if $\PL_\ell(x)=\{+1\}$, $-1$ if $\PL_\ell(x) = \{-1\}$, $0$ if $\PL_\ell(x) = \{\}$, and $!$ if $\PL_\ell(x) = \{-1,+1\}$ (Eqn~(\ref{eq:provisional-label})). 
%The interpretation of $\yh_\ell(x) \in \{+1,-1,0\}$ is given in Definition~\ref{def:accurate-bias-estimate}.

The label $\yh_\ell(x) = \, !$ indicates that at level $\ell$, there is conflicting evidence about $x$; it is a ``known unknown''~\citep{R11}. Our active learning algorithm makes its \emph{focused queries} at level $\ell+1$ in the vicinity of such points $x$, that is, in $\B_{\ell+1}(x)$.

% to be $-1$ if $\PL_\ell(x) = %\{-1\}$, or $+1$ if $\PL_\ell(x) = \{+1\}$, or $0$ if $\PL_\ell(x) = \{\}$, or $!$ %if $\PL_\ell(x) = \{-1,+1\}$, %and remains fixed thereafter. Points labeled $!$ are ``known unknowns'': they are %near the decision boundary and %need further investigation. Points labeled $0$ show no significant bias.

@@

The active learning algorithm, shown in Figs~\ref{alg:main} and \ref{alg:sampling}, has two types of query. \emph{Background queries} are random draws from $X$ and correspond to passive learning. \emph{Focused queries} are made in the vicinity of ``uncertain'' points and correspond to active learning. The \emph{uncertainty region} $U_\ell$ at level $\ell \geq 1$ consists of points $x \in X$ on which the prediction from the previous level is ``known unknown'', $\yh_{\ell-1}(x) = \,!$. Focused queries are drawn at random from balls in $\B_\ell$ containing such points.

On each iteration of the main loop, (at most) one focused query is made as well as a background query. The focused query comes from the lowest-numbered uncertainty region $U_\ell$ that is nonempty.

Once these queries are made, bias estimates $\yh(B)$ are updated, along with label-sets $\PL_\ell(x)$, label-estimates $\yh_\ell(x)$, and finally the uncertainty regions $U_\ell$. Then the next iteration begins.

The querying process can be stopped at any time, whereupon labels are assigned as follows:
\begin{equation}
\yh(x) = 
\left\{
\begin{array}{cl}
\yh_\ell(x) & \mbox{largest $\ell$ with $\yh_\ell(x) \in \{-1,+1\}$, if such an $\ell$ exists} \\
0 & \mbox{``don't know'', otherwise}
\end{array}
\right.
\label{eq:final-label}
\end{equation}

\begin{figure}[t]
\framebox{
\begin{minipage}[t]{6in}
\begin{center}
\begin{minipage}[t]{6in}
\vspace{.05in}

Initialize uncertainty regions at all levels: $U_0 = X$ and $U_\ell = \emptyset$ for $\ell \geq 1$

\vspace{.1in}

Initialize labels at all levels to ``unavailable'': \ $\yh_\ell(x) = \bot$ for all $x$ and $\ell \geq 0$

\vspace{.1in}

Repeat:
\begin{itemize}[leftmargin=0.5cm]
\item If there is a level $\ell \geq 0$ such that $U_\ell \neq \emptyset$:
\begin{itemize}[leftmargin=0.25cm]
\item For the smallest such level $\ell'$, run {\tt Focused-query}($\ell', U_{\ell'}$) // Fig.~\ref{alg:sampling}
\end{itemize}
\item {\tt Background-query} // Fig.~\ref{alg:sampling}
\item Update labels:
\begin{itemize}[leftmargin=0.25cm]
\item Update bias-estimates $\yh(B)$ // Fig.~\ref{alg:bias-estimate}
\item For each $x \in X$ and level $\ell \geq 0$ for which all $\{\yh(B): B \in \B_\ell(x)\}$ are available:
\begin{align}
\PL_\ell(x) &= \{s \in \{-1, +1\}: \ \mbox{there exists $B \in \B_{\leq \ell}(x)$ s.t. $\yh(B) = s$ and} \notag \\ 
             &                       \hspace{1.5in}\mbox{no $B' \in \B_{\leq \ell}(x)$ has $X_{B'} \subsetneq X_B$.} \} \label{eq:PL} \\ 
\yh_\ell(x) &= 
\left\{
\begin{array}{cl}
+1 & \mbox{if $\PL_\ell(x) = \{+1\}$} \\
-1 &  \mbox{if $\PL_\ell(x) = \{-1\}$} \\
0 & \mbox{if $\PL_\ell(x) = \{\}$} \\
! & \mbox{if $\PL_\ell(x) = \{-1,+1\}$}
\end{array}
\right.
\label{eq:provisional-label}
\end{align}
